\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography \\ or \\ Charting antigenic drift and evolutionary success of seasonal influenza lineages

### 2013-01-15 flux.tex
Clarify section on model comparison.

We test model performance by constructing training datasets representing 90\% of the HI measurements in each of the four datasets and test datasets representing the remaining 10\% of the measurements in each dataset. 
By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare these predicted titers to observed titers in the test dataset. 
We find that a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations and expected titer is relative to the maximum titer of a particular ferret serum, performs well, yielding an average absolute predictive error of 0.91 log$_2$ HI titers for A/H3N2 (Table \ref{errortable}).
We find that model with two dimensional antigenic locations predict test data to the same precision or slightly better precision as 3D models, but perform substantially better than 1D models or models with four or more dimensions (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
The finding of a low-dimensional map across influenza lineages extends previous studies in A/H3N2 \cite{Smith04} and remains an interesting empirical observation.
We extend this 2D model by estimating the strength of overall reactivity of each serum rather than fixing this at its maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
We refer to these estimates as serum effects and virus effects, respectively.
We find that including these effects decreases test error further, ranging from an improvement of 0.03 in A/H1N1 to an improvement of 0.14 in A/H3N2 (Table \ref{errortable}).

%%% errortable %%%
\begin{table}[h]
	\centering
	\caption{\textbf{Absolute prediction error of log$_2$ HI titer for test data across models and datasets.}}
	\label{errortable}		
	\begin{tabular}{ c c c c c c c c } 
	\hline
	\multicolumn{4}{c}{Model} 	&	\multicolumn{4}{c}{Test error}  \\
	Dimen 	& 	Serum effects 	&	Virus effects	& Covariance	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
	\hline	
	1D 		&	Fixed 			&	None			& None			&	1.07	&	0.92	&	0.94	&	1.08	\\	
	2D 		&	Fixed 			&	None			& None			&	0.91 	&	0.74 	&	0.77	&	0.83	\\
	3D 		&	Fixed 			&	None			& None			&	0.91	&	0.74 	&	0.79	&	0.84	\\
	4D 		&	Fixed 			&	None			& None			&	0.95	&	0.75 	&	0.81	&	0.86	\\
	5D 		&	Fixed 			&	None			& None			&	0.99	&	0.76 	&	0.84	&	0.88	\\	
	2D 		&	Estimated 		&	None			& None			&	0.79	&	0.74 	&	0.68	&	0.76	\\	
	2D 		&	Estimated 		&	Estimated		& None			&	0.77	&	0.71 	&	0.65	&	0.73	\\	
	2D 		&	Estimated 		&	Estimated		& Diffusion		&	0.76	&	0.71 	&	0.64	&	0.72	\\		
	\hline
	\end{tabular}	
\end{table}

Previous work on influenza antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}. 
Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of phylogenetic tree according to a Brownian motion process (see Methods).
The diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process, and find a small increase in predictive accuracy when including covariance induced by the diffusion (Table \ref{errortable}).
Additionally, modeling antigenic diffusion gives the advantage of directly correlating evolutionary outcomes with changes in antigenic phenotype.

@@

\subsection*{Posterior samples and summaries}

>>>>

We test model performance by constructing training datasets representing 90\% of the HI measurements for each of the four influenza lineages and test datasets representing the remaining 10\% of the measurements for each lineage. 
By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare these predicted titers to observed titers. 
We begin with a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations, expected titer is relative to the maximum titer of a particular ferret serum and virus and serum locations follow a flat prior.
We find that this model performs well, yielding an average absolute predictive error of between 0.79 and 0.95 log$_2$ HI titers across influenza lineages (model 1; Table \ref{errortable}).
We extend this model to include a prior on virus and serum locations that assumes antigenic location along dimension 1 increases in proportion to date of virus or serum sampling, resulting in a small improvement in predictive error (model 2; Table \ref{errortable}).

%%% errortable %%%
\begin{table}[h]
	\centering
	\caption{\textbf{Average absolute prediction error of log$_2$ HI titer for test data across models and datasets.}}
	\label{errortable}	
	\makebox[\textwidth][c]{
	\begin{tabular}{ c c c c c c c c } 
	\hline
	\multicolumn{4}{c}{} 	&	\multicolumn{4}{c}{Test error}  \\
	Model	&	Dimen 	& 	Serum/virus effects 	& Location Prior					& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
	\hline		
	1		&	2D 		&	Fixed $\ses$ 			& Flat								&	0.95 	&	0.79 	&	0.80	&	0.93	\\
	2		&	2D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.91 	&	0.74 	&	0.77	&	0.83	\\	
	3		&	1D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	1.07	&	0.92	&	0.94	&	1.08	\\		
	4		&	3D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.91	&	0.74 	&	0.79	&	0.84	\\
	5		&	4D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.95	&	0.75 	&	0.81	&	0.86	\\
	6		&	5D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.99	&	0.76 	&	0.84	&	0.88	\\				
	7		&	2D 		&	Estimated $\ses, \ves$	& $\beta,\virussd,\serumsd$			&	0.77	&	0.71 	&	0.65	&	0.73	\\	
	8		&	2D 		&	Estimated $\ses, \ves$	& $\beta,\virussd,\serumsd,\tree$	&	0.76	&	0.71 	&	0.64	&	0.72	\\		
	\hline
	\end{tabular}	
	}
\end{table}

We find that decreasing dimensionality to a single antigenic dimension results in a substantially less predictive model (model 3; Table \ref{errortable}), and that increasing dimensionality to three dimensions is equal or slightly worse in predictive power (model 4; Table \ref{errortable}).
Increasing dimensionality further to four or five dimensions results in continued declines in test error (models 5 and 6; Table \ref{errortable}).
Consequently, we specify a two dimensional model in all subsequent analyses.
The finding of a low-dimensional map across influenza lineages extends previous studies in A/H3N2 \cite{Smith04} and remains an interesting empirical observation.
We extend the 2D model by estimating the strength of overall reactivity of each serum rather than fixing this at its maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
We refer to these estimates as serum effects and virus effects, respectively.
We find that including these effects decreases test error further, ranging from an improvement of 0.03 in A/H1N1 to an improvement of 0.14 in A/H3N2 (model 7; Table \ref{errortable}).

Previous work on influenza antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}. 
Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of phylogenetic tree according to a Brownian motion process (see Methods).
The diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process, and find a small increase in predictive accuracy when including covariance induced by the diffusion (model 8; Table \ref{errortable}).
Additionally, modeling antigenic diffusion has the advantage of allowing direct correlation of evolutionary outcomes with changes in antigenic phenotype.

@@

\subsection*{Posterior samples}
